The status of the microbial census: an update

### 2015-11-09 doc/notebook/Archaeal_data_acquisition.Rmd
Use github flavored markdown

* Go to search window
* Set Search_Fields to "ARB_color" and use "1"; click on the equal sign to make it not equal
* Hit Mark Listed, Unmark Rest button (N=1493493)

@@

* ARB_color != 1 & tax_rdp == "*Archaea*" (N=40288)
* ARB_color != 1 & tax_greengenes == "*Archaea*" (N=19795)
* ARB_color != 1 & tax_slv == "*Archaea*" (N=17641)
* ARB_color != 1 & tax_embl == "*Archaea*" (N=44574)
* ARB_color != 1 & (tax_rdp | tax_greengenes | tax_embl | tax_slv) (N=46387 )

Let's stick with the archaeal sequences that have an RDP taxonomy and we will
analyze them in RDP space, but we can always go back to the taxonomy provided by
the other systems if we'd like. A quick spot checking suggests that the RDP
taxonomy information is actually richer than that of the EBML taxonomy
information.

* Return to Search and Query window
* Change search field to "ARB_color", enter "1", and click the equal sign to be not-equal
* Change the second search field to "tax_rdp" and set it to "*Archaea*"
* Hit "Search"
* Hit Mark Listed Unmark Rest (N=40,288)
* Click "Write to Fields of Listed", select the "remark" field, and enter "good archaea"
* Click "Write"
* In main ARB window go File->Export->Export to external format
* Select Compress -> "Vertical Gaps" and fasta_mothur.eft as the format
* Rename `noname.fasta` to `archaea.fasta`

@@

* Go Tree -> NDS
* Click "name", "acc", "tax_rdp". The "tax_rdp" field should have 250 characters
* Unclick everything else
* Click "Close"
* Go File->Export->Export fields
* Set the file name to "archaea.taxonomy" and Column output to "TAB separated"
* Click "SAVE"

@@

The next thing we need to do is to clean up the archaea.taxonomy.nds file
to make it into a proper, mothur compatible archaea.taxonomy file. First we
want to remove the names of any sub taxa (e.g. suborder). To do this we need
to get the RDP list of names and the taxonomic level they belong to...

```{r get_rdp_names, engine="bash"}
wget -N http://sourceforge.net/projects/rdp-classifier/files/RDP_Classifier_TrainingData/RDPClassifier_16S_trainsetNo10_rawtrainingdata.zip
unzip -o RDPClassifier_16S_trainsetNo10_rawtrainingdata.zip
mv RDPClassifier_16S_trainsetNo10_rawtrainingdata/* data/references/
rm -rf RDPClassifier_16S_trainsetNo10_rawtrainingdata*
```

Now we're read to do some R'ing to get the taxonomy file formatted properly...

@@

taxonomy <- gsub(" ", "_", tax_data$V3)  #convert any spaces to underscores
taxonomy <- gsub("[^;]*_incertae_sedis$", "", taxonomy) #remove the unknowns
taxonomy <- gsub('\"', '', taxonomy) #remove quote marks

#now we need to pull out the taxon names for the sub taxa levels
levels <- read.table(file="data/references/trainset10_db_taxid.txt", sep="*", stringsAsFactors=F)
subs <- levels[grep("sub", levels$V5),] #get the sub taxa names
sub.names <- subs$V2

tax.split <- strsplit(taxonomy, split=";")  #split the taxonomy string by semicolon

#function to see which taxa names in a list are found in the sub taxa list
remove.subs <- function(tax.vector){
    return(tax.vector[which(!tax.vector %in% sub.names)])
}

#apply sub taxa finding function to all sequences
no.subs <- lapply(tax.split, remove.subs)

#merge data together by semicolon for each sequence
no.subs.str <- unlist(lapply(no.subs, paste, collapse=";"))

#Tack a semicolon on to the end of each sequence
no.subs.str <- paste0(no.subs.str, ";")

write.table(cbind(seq_names, no.subs.str), "data/mothur/archaea.tax", row.names=F,
                                                col.names=F, quote=F, sep="\t")
```

@@

```
            Start    End	NBases	Ambigs	Polymer	NumSeqs
Minimum:	1        4058	900	    0    	4	1
2.5%-tile:	357      4692	905	    0    	5	1008
25%-tile:	365      4783	935	    0   	5	10073
Median: 	379      7409	1262	0   	5	20145
75%-tile:	456      7729	1349	0    	6	30217
97.5%-tile:	1746	 7984	1470	4   	7	39281
Maximum:	2321	 8711	2174	46     	32	40288
Mean:	    513.659	 6698.2	1185.3	0.3	    5.44244
# of Seqs:	40288
```

@@

```{r find_screening_points}
data <- read.table(file="data/mothur/archaea.summary", header=T, row.names=1)
quantile(data$start, probs=seq(0.9,1,0.01))
start <- 2000

quantile(data$end, probs=seq(0,0.1,0.01))
end <- 4058

@@

```{r screen_data, engine="bash"}
mothur "#set.current(processors=8);
        screen.seqs(fasta=archaea.fasta, taxonomy=archaea.tax, start=2000, end=4058, maxambig=2, maxhomop=8, minlength=900, inputdir=data/mothur/, outputdir=data/mothur/);"
```

@@

```{r single_cell, engine="bash"}
mothur "#set.dir(input=data/mothur, output=data/mothur);
        align.seqs(fasta=single_cell.archaea.fasta, reference=archaea.good.fasta);
        screen.seqs(fasta=current, start=2000, end=4058, maxambig=2, maxhomop=8, minlength=900);
        classify.seqs(fasta=current, reference=data/references/trainset10_082014.pds.fasta, taxonomy=data/references/trainset10_082014.pds.tax, cutoff=80, processors=8)"
```

@@

if(length(extra) == 0){
    write.table(extra, file="missing.txt", row.names=F, col.names=F, quote=FALSE,  sep="\t")
}

>>>>

* Go to search window
* Set Search_Fields to "ARB_color" and use "1"; click on the equal sign to make it not equal
* Hit Mark Listed, Unmark Rest button (N=1652440)

@@

* ARB_color != 1 & tax_rdp == "*Archaea*" (N=46146)
* ARB_color != 1 & tax_greengenes == "*Archaea*" (N=19830)
* ARB_color != 1 & tax_slv == "*Archaea*" (N=59240)
* ARB_color != 1 & tax_embl == "*Archaea*" (N=57863)

Let's stick with the archaeal sequences that have an Silva taxonomy and we will
analyze them in Silva space, but we can always go back to the taxonomy provided by
the other systems if we'd like.

* Return to Search and Query window
* Change search field to "ARB_color", enter "1", and click the equal sign to be not-equal
* Change the second search field to "tax_silva" and set it to "*Archaea*"
* Hit "Search"
* Hit Mark Listed Unmark Rest (N=59,240)
* Click "Write to Fields of Listed", select the "remark" field, and enter "good archaea"
* Click "Write"
* In main ARB window go File->Export->Export to external format
* Select Compress -> "Vertical Gaps" and fasta_mothur.eft as the format
* Rename `noname.fasta` to `data/mothur/archaea.fasta`

@@

* Go Tree -> NDS
* Click "name", "acc", "tax_slv". The "tax_slv" field should have 250 characters
* Unclick everything else
* Click "Close"
* Go File->Export->Export fields
* Set the file name to "archaea.taxonomy" and Column output to "TAB separated"
* Click "SAVE"

@@

The next thing we need to do is to clean up the archaea.taxonomy.nds file
to make it into a proper, mothur compatible archaea.taxonomy file. Now we're
ready to do some R'ing to get the taxonomy file formatted properly...

@@

taxonomy <- gsub(" ", "_", tax_data$V3)  #convert any spaces to underscores
taxonomy <- gsub('\"', '', taxonomy) #remove quote marks

write.table(cbind(seq_names, taxonomy), "data/mothur/archaea.tax", row.names=F,
                                                col.names=F, quote=F, sep="\t")
```

@@

```
			Start	End		NBases	Ambigs	Polymer	NumSeqs
Minimum:	1		6806	900		0		4		1
2.5%-tile:	1004	7718	905		0		5		1482
25%-tile:	1006	8794	943		0		5		14811
Median: 	1039	11168	1308	0		5		29621
75%-tile:	1056	11747	1398	0		6		44431
97.5%-tile:	3020	11909	1469	3		8		57760
Maximum:	5159	12877	2962	46		25		59240
Mean:		1270.65	10393.5	1204.49	0.26683	5.53047
# of Seqs:	59240
```

@@

```{r find_screening_points}
data <- read.table(file="data/mothur/archaea.summary", header=T, row.names=1)
quantile(data$start, probs=seq(0.9,1,0.01))
start <- 3000

quantile(data$end, probs=seq(0,0.1,0.01))
end <- 7800

@@

```{r screen_data, engine="bash"}
mothur "#set.current(processors=8);
        screen.seqs(fasta=archaea.fasta, taxonomy=archaea.tax, start=3000, end=7800, maxambig=2, maxhomop=8, minlength=900, inputdir=data/mothur/, outputdir=data/mothur/);"
```

@@

```{r single_cell, engine="bash"}
mothur "#set.dir(input=data/mothur, output=data/mothur);
        align.seqs(fasta=single_cell.archaea.fasta, reference=archaea.good.fasta);
        screen.seqs(fasta=current, start=2000, end=4058, maxambig=2, maxhomop=8, minlength=900);
        classify.seqs(fasta=current, reference=archaea.good.fasta, taxonomy=archaea.good.tax, cutoff=80, processors=8)"
```

@@

if(length(extra) != 0){
    write.table(extra, file="missing.txt", row.names=F, col.names=F, quote=FALSE,  sep="\t")
}
